---
title: "Non-Linear Least Squares"
description: |
  Using Non-Linar Least Squares (NLS) analysis on lizards sampled by pitfall traps at the Jornada Basin LTER site to estimate parameters of a length to weight model for lizard popilations. 
author:
  - name: Devin Ngo
    url: {}
date: 2022-03-13
output:
  distill::distill_article:
    self_contained: false
    code_folding: Show Code
---
## Overview

This data set contains data on lizards sampled by pitfall traps at 11 consumer plots at Jornada Basin LTER site from 1989 - 2006. This study was conducted to observe how shifts affect the spatial and temporal availability of resources for consumers. The variables measured in this data include species, sex, snout-vent length, total length, and weight. This report uses non-linear least squares (NLS) to estimate parameters of a length to weight model for lizard populations in New Mexico. NLS will be used to look at a model for all species and then for a model for male western whiptail lizards. These models will be compared to see which one is a better fit in predicting lizard weight as a function of snout-to-vent length. 

## Setup
```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

```{r class.source = 'fold-show', warning = FALSE}
# Attached packages:
library(tidyverse)
library(here)
library(Metrics)
library(purrr)
library(janitor)
library(kableExtra)
library(broom)
library(modelr)
```

## Data Wrangling 

Reading in and wrangling lizards data, as well as adding new columns for log of weight and log of snout to vent length:  

```{r}
# Reading in data
lizards <- read_csv(here("_posts", "2022-03-13-nls", "data", "lizard.csv")) %>% 
  clean_names() %>% 
  mutate(sex = case_when(
    sex == "M" ~ "Male",
    sex == "F" ~ "Female")) %>% 
    mutate(log_weight = log(weight)) %>% #Adding log of weight column 
  mutate(log_length = log(sv_length)) %>%  #Adding log of sv_length column 
  drop_na()
```

## Model Fit Selection

Introducing a snout length to weight model of the following form to the lizards data frame:

\[
W = a(SVL)^b
\]

W is weight (g), SVL is snout to vent length (mm), and a and b are the parameters that need to be fitted. 

```{r}
# Log transforming the model
my_guess_model <- lm(log_weight ~ log_length, data = lizards) 
```

```{r}
# Supplying the NLS start list with regression coefficients 
coef <- coefficients(my_guess_model) 

# Creating and storing a function of the weight model
weight_model <- function(a, sv_length, b){
  weight = a*(sv_length)^b
  return(weight)
}
```

## NLS Model Results

```{r}
# Creating the NLS model 
lizard_nls=nls(weight ~ weight_model (a,sv_length,b),
               data = lizards,
               start = list(b = coef[2], 
                            a = exp(coef[1])))

# NLS model results in a table
nls_tidy <- tidy(lizard_nls) %>% 
  select(-statistic) %>% 
  mutate(p.value = case_when(
    p.value < 0.001 ~ "<0.001")) %>% 
  kable(col.names = c("Parameter", "Coefficient", "Standard Error", "P-Value"),
        caption = "Table 1. NLS analysis results for all lizard species. Data includes coefficients for the parameters, standard error, and the p-value.", 
        digits = 4) %>% 
  kable_styling(full_width = FALSE)

nls_tidy
```

## Visualizing the Model Prediction Based on NLS
```{r}
# To create model prediction based on NLS
lizard_augment <- augment(lizard_nls)

# Fitted model on a plot with female and male lizards separated by color 
ggplot() +
  geom_point(data = lizards, aes(
             x = sv_length,
             y = weight, 
             color = sex)) +
  geom_line(data = lizard_augment, aes(
            x = sv_length,
            y = .fitted,
            color = "Model Prediction"),
            size = 1) +
  scale_color_manual(values = c("lightslategray", "lightskyblue1", "firebrick")) +
  labs(x = "Snout to Vent Length (mm)", 
       y = "Weight (g)", 
       color = "Legend") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

***Figure 1.** Lizard body weight (g) versus snout to vent length (mm) of all lizard species, separated by sex. The red line represents the model prediction based on the NLS model. Data: Lightfoot & Whitford. (2020).*

## NLS Model - Male Western Whiptail Lizards

```{r}
# Filtering out the dataset for male Western Whiptail lizard (Cnemidophorus tigrisatus) 
male_whiptails <- lizards %>% 
  filter(spp %in% c("CNTI"), 
         sex %in% c("Male"))

# Log transforming the model for the subset
guess_male_whip <- lm(log_weight ~ log_length, data = male_whiptails)

# Pulling out regression coefficients 
coef_male_whip <- coefficients(guess_male_whip) 

# Fitting NLS model to the subset 
nls_male_whip <- nls(weight ~ weight_model(a, sv_length, b),
                     data = male_whiptails,
                     start = list(b = coef_male_whip[2], 
                                  a = exp(coef_male_whip[1]))) 

# NLS model results in a table
nls_male_whip_tidy <- tidy(nls_male_whip) %>% 
  select(-statistic) %>% 
  mutate(p.value = case_when(
    p.value < 0.001 ~ "<0.001",
    p.value > 0.001 ~ as.character(round(p.value, 4)))) %>% 
  kable(col.names = c("Parameter", "Coefficient", "Standard Error", "P-Value"),
        caption = "Table 2. NLS analysis results for male western whiptail lizards. Data includes coefficients for the parameters, standard error, and the p-value.",
        digits = 4) %>% 
  kable_styling(full_width = FALSE)

nls_male_whip_tidy 
```

## Visualizing and Comparing NLS Models

```{r}
# To create model prediction based on NLS for subset
male_whip_augment <- augment(nls_male_whip)

whiptails_general <- data.frame(lizards, lizard_augment) %>% 
  filter(spp %in% c("CNTI"), 
         sex %in% c("Male"))

# Comparing the output from the species specific nls model to the general nls model for all species by graphing the model fits on the male western whiptail data 
ggplot() +
  geom_point(data = male_whiptails, aes(
             x = sv_length,
             y = weight,
             color = "Data")) +
  geom_line(data = male_whip_augment, aes(
            x = sv_length,
            y = .fitted,
            color = "Male Whiptail Model"),
            size = 1) +
  geom_line(data = whiptails_general, aes(
            x = sv_length,
            y = .fitted,
            color = "General Lizards Model"),
            size = 1) +
  scale_color_manual(values = c("black", "aquamarine4", "deeppink4")) +
  theme_minimal() +
  labs(x = "Snout to Vent Length (mm)", 
       y = "Weight (g)") +
  theme(legend.title = element_blank(),
        legend.position = "bottom")

# Recording the RMSE for both models 
whip_rmse <- rmse(nls_male_whip, data = male_whiptails) 
whip_gen_rmse <- rmse(lizard_nls, data = male_whiptails) 
```

***Figure 2.** Lizard body weight (g) versus snout to vent length (mm) for male western whiptail lizards. The green line represents a general NLS model for all lizard species, while the red line represents the NLS model specifically trained for male western whiptail lizards. The RMSE for the general model is `r round(whip_gen_rmse, 3)` while for the male whiptail specific model, the RMSE is `r round(whip_rmse, 3)`. The lower RMSE score for the specific model indicates it is a better fit line for predicting data (in this case: weight as a function of snout to vent length). Data: Lightfoot & Whitford. (2020).*

## Data Citation
Lightfoot, D. and W.G. Whitford. 2020. Lizard pitfall trap data from 11 NPP study locations at the Jornada Basin LTER site, 1989-2006 ver 37. Environmental Data Initiative. https://doi.org/10.6073/pasta/4a6e258fb49c31e222ecbbcfd128967f
